---
title: "Group Project"
subtitle: "Project Proposal and Data"
author: "Ellie, Avery, and Bo"
bibliography: references.bib
nocite: |
  @*
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: false
  message: false
  warning: false


---

```{r setup}

library(knitr)
library(tidyverse)
library(bibtex)

healthcare_spending <- read_csv("data/government_health_spending_per_person_us.csv")
life_exp <- read_csv("data/lex.csv")
```
## Introduction



## The Data

We are exploring government health spending per person in US dollars and life expectancy at birth.
The life expectancy at birth dataset gives the number of years a newborn infant will live if the current mortality rates were to stay the same throughout their life. This data is from the Institute for Health Metrics and Evaluation from a study that was published in 2018 [@Life_Expectancy_at_Birth_Gapminder].
The government health spending per person dataset gives the average health cost per person that has been paid by the government during a given year, in US dollars using the average exchange rate. This data is comes from the World Health Organization's Global Health Observatory [@Global_Health_Expenditure_Database].

We assume there is a relatively strong positive relationship between government spending and life expectancy. We hypothesize this because if the government spends more money on healthcare for each individual person, they will hopefully be receiving more health care and should be expected to live longer [@Anwar_Hyder_Mohamed_Nor_Younis_2023].

### Data Cleaning

To start cleaning up our data we removed all years that did not have information on government health spending per person in US dollars. This left us with a data from 1995-2010 for 190 countries.

We also filtered for countries/regions only present in both datasets which excluded the Cook Islands, Niue, Hong Kong, Liechtenstein, North Korea, Palestine, South Sudan, and Taiwan. This may skew our outcomes, as many of these regions are either small in population or area, or have experienced political instability. 

```{r}

healthcare_spending_long <- pivot_longer(healthcare_spending, 
                                         `1995`:`2010`, 
                                         names_to = "Year",
                                         values_to = "HealthSpendingPerCapita") 

life_exp_long <- life_exp |>
  select(country, `1995`:`2010`) |>
  pivot_longer(`1995`:`2010`,
               names_to = "Year",
               values_to = "LifeExpAtBirth") |>
  filter(Year >= 1995 & Year <= 2010)

joined_data_long <- inner_join(healthcare_spending_long, 
                              life_exp_long, 
                              by = c("country", "Year"))
```

These restrictions meant that we had no missing values for life expectancy at birth. 
We had a total of 33 missing values from health spending per person, this table shows a few of those:

```{r}
joined_data_long |>
  filter(is.na(HealthSpendingPerCapita)) |>
  slice_sample(n = 5) |>
  kable()
```

The main countries that had missing values for health care spending per capita were Afghanistan from 1995-2001 as well as Zimbabwe and Somalia from 2002-2010. We believe that these numbers are missing due to ongoing conflict in these countries during those time periods. 
There were a couple of other countries with missing values for health care spending per capita, however, they appear to be missing at random since they are one-offs.
